Isolator: accurate and stable analysis of isoform-level expression in RNA-Seq experiments

### 2016-11-30 paper.tex
Nix Katze
tweak some terminology

\title{Isolator: accurate and consistent analysis of isoform-level expression in RNA-Seq experiments}

\author[1]{Daniel C. Jones}
\author[2,6]{Kavitha T. Kuppusamy}
\author[3]{Nathan J. Palpant}
\author[4]{Xinxia Peng}
\author[4]{Michael G. Katze}
\author[2,5]{Charles E. Murry}
\author[2,6]{Hannele Ruohola-Baker}
\author[1,7,8]{Walter L. Ruzzo}

@@

\begin{abstract}
    While RNA-Seq has enabled great progress towards the goal of wide-scale
    isoform-level mRNA quantification, short reads have limitations when
    resolving complex or similar sets of isoforms. As a result, estimates of
    isoform abundance carry far more uncertainty than those made at the gene
    level. When confronted with this uncertainty, commonly used methods produce
    estimates that are often high-variance---small perturbations in the data
    often produce dramatically different results, confounding downstream
    analysis. We introduce a new method, Isolator, which analyzes all samples in
    an experiment in unison using a simple Bayesian hierarchical model. Combined
    with aggressive bias correction, it produces estimates that are
    simultaneously accurate and stable. In a comprehensive comparison of
    accuracy and variance, we show that this property is unique to Isolator. We
    further demonstrate that the approach of modeling an entire experiment
    enables new analyses, which we demonstrate by examining splicing
    monotonicity across several time points in the development of human
    cardiomyocyte cells.
\end{abstract}

@@

The four SEQC samples consist of two commercially available reference RNA
samples, labeled $A$ and $B$, and two synthetic samples formed by mixing $A$ and
$B$ in specific proportions. Sample $C$ is composed of 75\% $A$ and 25\% $B$,
while sample $D$ is 25\% $A$ and 75\% $B$. If we estimate transcript abundance
for each sample, producing estimates $a, b, c, d$, then for these estimates to
be in agreement (``coherent'', in our terminology) we should expect $c \appprox
0.25a + 0.75b$, and similarly $d \approx 0.75a + 0.25b$. In other words, mixing
RNA then estimating transcript abundance should be approximated by estimating
transcript abundance then mixing estimates.

@@

Results from RSEM are particularly informative. We generated maximum likelihood
and posterior mean estimates using the same aligned reads. Merely switching the
estimate from the former to the latter increased the correlation by
approximately 5\%. This stands in contrast to comparisons to qPCR which suggest
reduced accuracy of posterior mean estimates, demonstrating that consistency and
accuracy, while related, are two separate axis of comparison.

>>>>

\title{Isolator: accurate and stable analysis of isoform-level expression in RNA-Seq experiments}

\author[1]{Daniel C. Jones}
\author[2,6]{Kavitha T. Kuppusamy}
\author[3]{Nathan J. Palpant}
\author[4]{Xinxia Peng}
%\author[4]{Michael G. Katze}
\author[2,5]{Charles E. Murry}
\author[2,6]{Hannele Ruohola-Baker}
\author[1,7,8]{Walter L. Ruzzo}

@@

\begin{abstract}
    While RNA-Seq has enabled great progress towards the goal of wide-scale
    isoform-level mRNA quantification, short reads have limitations when
    resolving complex or similar sets of isoforms. As a result, estimates of
    isoform abundance carry far more uncertainty than those made at the gene
    level. When confronted with this uncertainty, commonly used methods produce
    estimates that are often high-variance---small perturbations in the data
    often produce dramatically different results, confounding downstream
    analysis. We introduce a new method, Isolator, which analyzes all samples in
    an experiment in unison using a simple Bayesian hierarchical model. Combined
    with aggressive bias correction, it produces estimates that are
    simultaneously accurate and show high agreement between samples. In a
    comprehensive comparison of accuracy and variance, we show that this
    property is unique to Isolator. We further demonstrate that the approach of
    modeling an entire experiment enables new analyses, which we demonstrate by
    examining splicing monotonicity across several time points in the
    development of human cardiomyocyte cells.
\end{abstract}

@@

The four SEQC samples consist of two commercially available reference RNA
samples, labeled $A$ and $B$, and two synthetic samples formed by mixing $A$ and
$B$ in specific proportions. Sample $C$ is composed of 75\% $A$ and 25\% $B$,
while sample $D$ is 25\% $A$ and 75\% $B$. If we estimate transcript abundance
for each sample, producing estimates $a, b, c, d$, then for these estimates to
be in agreement (``coherent'', in our terminology) we should expect $c \approx
0.25a + 0.75b$, and similarly $d \approx 0.75a + 0.25b$. In other words, mixing
RNA then estimating transcript abundance should be approximated by estimating
transcript abundance then mixing estimates.

@@

Results from RSEM are particularly informative. We generated maximum likelihood
and posterior mean estimates using the same aligned reads. Merely switching the
estimate from the former to the latter increased the correlation by
approximately 5\%. This stands in contrast to comparisons to qPCR which suggest
reduced accuracy of posterior mean estimates, demonstrating that coherence and
accuracy, while related, are two separate axis of comparison.
